Iterative Reduction: Parsing and Interpreting Code Without Trees

### 2025-01-11 Paper/paper.tex
Added some formal languages to the paper

@@

>>>>

\begin{theorem} \label{thrm:complexity}

@@

\section{$\beta$-Reducible Languages}

In this section we will try to understand better the class of languages which can be parsed using $\beta$-reduction.
The definition of $\beta$-reduction in this section will differ slightly from the remainder of this paper, but core ideas will be preserved.
We will construct a hierarchy of $\beta$-reducible languages, which we will compare with the Chomsky Hierarchy.

\subsection{$\beta$-Reducible Languages}

The simplest class in our hierarchy will be $\beta$-reducible languages, the class of which we will denote by \BR.
Define a \BR-schema to be a tuple $B=(\Sigma,\Gamma,s,\hb)$ where
\begin{enumerate}
    \item $\Sigma$ is a finite alphabet of terminals,
    \item $\Gamma$ is a finite alphabet of non-terminals (so $\Sigma$ and $\Gamma$ are disjoint),
    \item $s$ is a state, which is a map $\Sigma\to\Gamma$,
    \item $\hb$ is an initial $\beta$-reducer, which is a partial function $\hb\colon\Gamma\times(\Gamma\cup\set{\epsilon,\square})\partial\Gamma\cup\set{\epsilon,\bullet}$.
\end{enumerate}
Given a \BR-schema, we define a $\beta$-reducer to be a function $\beta\colon(\Gamma\cup\set\bullet)^*\longto(\Gamma\cup\set\bullet)^*$ as follows: given a string $\xi\in(\Gamma\cup\set\bullet)^*$ we define $\beta(\xi)$ as follows:
\begin{enumerate}
    \item If $\xi=\bullet$, then $\beta(\xi)=\xi$
    \item If $\xi=\sigma\in\Gamma$, then $\beta(\xi)=\hb(\sigma,\square)$ if it is defined, and $\sigma$ otherwise,
    \item If $\xi=\sigma_1\sigma_2\xi'$, then
        \begin{enumerate}
            \item $\beta(\xi)=\hb(\sigma_1,\sigma_2)\xi'$ if it is defined, otherwise
            \item $\beta(\xi)=\hb(\sigma_1,\epsilon)\sigma_2\xi'$ if it is defined, otherwise
            \item $\beta(\xi)=\sigma_1\beta(\sigma_2\xi')$.
        \end{enumerate}
\end{enumerate}
Then we define $\beta^*(\xi)$ to be the fixed point of successive applications of $\beta$ to $\xi$, i.e. $\beta^*(\xi)=\beta^n(\xi)$ where $\beta^n(\xi)=\beta^{n+1}(\xi)$.
For a word $w=\sigma_1\cdots\sigma_n=\in\Sigma^*$, we define $s(w)=s(\sigma_1)\cdots s(\sigma_n)\in\Gamma^*$.
The language \textit{generated} (or \textit{accepted}) by a \BR-schema $B$ is defined to be all the words $w\in\Sigma^*$ such that $\beta^*(s(w))=\bullet$.

Finally, let $\sigma\in\Gamma$: define the set $C(\sigma)=\bigcup_{n\geq1}C_n(\sigma)$ where $C_0(\sigma)=\set\sigma$ and $C_{n+1}(\sigma)=\set{\hb(\tau,\epsilon),\hb(\tau,\square)}[\tau\in C_n(\sigma)]$.
Call a \BR-schema \textit{valid} if $\sigma\notin C(\sigma)$ for all $\sigma\in\Gamma$.
A $\beta$-reducible language is one generated by a valid \BR-schema; the class of all $\beta$-reducible languages is denoted \BR.
The reason for requiring a \BR-schema to be valid is to ensure that the $\beta$-reducer halts.
Otherwise, suppose $\sigma\in C(\sigma)$, so we could have that $\hb(\sigma,\square)=\sigma$ for example, and then $\beta$ would never halt on the string $\sigma$.
Validity is not a strong requirement.

According to theorem \ref{thrm:complexity}, all $\beta$-reducible languages can be reduced by their $\beta$-reducer in $O(n)$ time.

Note that the empty word cannot be accepted by a \BR-schema, but we don't really lose any expressive power with this: we can always add another field to the \BR-schema saying whether or not it should accept the empty word.

For example, $\set{a^nb^n}[n\geq1]$ is in \BR.
We define $B=(\Sigma,\Gamma,s,\hb)$ as follows:
\begin{enumerate}
    \item $\Sigma=\set{a,b}$,
    \item $\Gamma=\set{S_a,S_b,S_{ab},S_{aab}}$,
    \item $s(a)=S_a$ and $s(b)=S_b$,
    \item $\hb$ is defined by the rules:
        \[ S_a,S_b\to S_{ab},\quad S_a,S_{ab}\to S_{aab},\quad S_{aab},S_b=S_{ab},\quad S_{ab},\square\to\bullet \]
\end{enumerate}
Indeed, a word of the form $a\cdots ab\cdots b$ will be mapped to $S_a\cdots S_aS_b\cdots S_b$, which will then be reduced to $S_a\cdots S_aS_{ab}S_b\cdots S_b\to S_a\cdots S_{aab}S_b\cdots S_b\to S_a\cdots S_{ab}\cdots S_b$, and so on until $S_{ab}$, which is then reduced
to $\bullet$.

Now, we claim that every regular language is in \BR{} (we will not consider languages containing the empty string).
Recall that regular languages are precisely the languages generated by regular expressions.
A regular expression over $\Sigma$ is a string defined by
\[ \c R \ccoloneqq \Sigma \gbar (\c R+\c R) \gbar (\c R\cdot\c R) \gbar \c R^+ \]
The language generated by $r\in\c R$ is $\c L(r)$, defined recursively:
\begin{enumerate}
    \item If $r=\sigma\in\Sigma$, then $\c L(r)=\set\sigma$,
    \item If $r=(r_1+r_2)$, then $\c L(r)=\c L(r_1)\cup\c L(r_2)$,
    \item If $r=(r_1\cdot r_2)$, then $\c L(r)=\c L(r_1)\cdot\c L(r_2)$,
    \item If $r=r_1^+$, then $\c L(r)=\c L(r_1)^+$ (where $L^+=\bigcup_{n\geq1}L^n$).
\end{enumerate}
(Importantly, these generate all the regular languages not containing the empty string.)

    A \BR-schema $B=(\Sigma,\Gamma,s,\hb)$ is in \emph{regular form} if
    \begin{enumerate}
        \item $\Gamma$ consists only of symbols of the form $S_\omega$ where $\omega$ is some string (in some alphabet).
        \item $s$ maps $\sigma$ to $S_\sigma$ for all $\sigma\in\Sigma$.
        \item $\hb$ has reduction rules only of the form $S_\omega,S_\sigma\to S_{\omega'}$ and $S_\omega,\square\to\bullet$.
    \end{enumerate}

    Every regular language can be generated by a \BR-schema in regular form.

    We proceed by induction on the structure of a regular expression.
    For singletons, this is trivial: $\set\sigma$ can be generated by the reduction rule $S_\sigma,\square\to\bullet$.

    Now if the regular expression is of the form $r_1+r_2$, by induction there exists \BR-schemas $B_1=(\Sigma,\Gamma_1,s,\hb_1)$ and $B_2=(\Sigma,\Gamma_2,s,\hb_2)$ for $\c L(r_1),\c L(r_2)$ respectively.
    We define $B=(\Sigma,\Gamma,s,\hb)$ as follows ($\Gamma$ is defined implicitly as all the symbols occurring in the following definition of $\hb$): we start by setting $\hb$ to be the union of $\hb_1$ and $\hb_2$.
    Furthermore, for every reduction rule $S_\omega,S\to S_{\omega'}$ in $\hb_i$, we add $S_{\mbox\omega_i},S\to S_{\mbox{\omega'}_i}$ to $\hb$.
    We are assuming that no such symbols are already in use; if they are we can just change the number in the subscript of $\mbox{\omega}_i$.

    If there exists symbols $S_\omega,S_\sigma$ such that $\hb_1(S_\omega,S_\sigma)=S_{\omega_1}$ and $\hb_2(S_\omega,S_\sigma)=S_{\omega_2}$ where $S_{\omega_1}\neq S_{\omega_2}$ then we will set $\hb(S_\omega,S_\sigma)=S_{\omega_1+\omega_2}$.
    Now, for any symbol of the form $S_{\omega_1+\omega_2}$ that we have just added, we add the following reduction rules: for any $\sigma\in\Sigma$,
    \begin{enumerate}
        \item if $\hb_i(S_{\omega_i},S_\sigma)=S_{\omega_i'}$ is defined for only one $i\in\set{1,2}$, then $\hb(S_{\omega_1+\omega_2},S_\sigma)=S_{\mbox{\omega_i'}_i}$,
        \item if $\hb_i(S_{\omega_i},S_\sigma)=S_{\omega_i'}$ is defined for both $i$s, and $\omega_1'=\omega_2'$, then $\hb(S_{\omega_1+\omega_2},S_\sigma)=S_{\omega_1'}=S_{\omega_2'}$,
        \item otherwise $\hb(S_{\omega_1+\omega_2},S_\sigma)=S_{\omega_1'+\omega_2'}$.
    \end{enumerate}

    We continue to regular expressions of the form $r_1\cdot r_2$.
    Again, we have \BR-schemas $B_1$ and $B_2$ which generate the languages for $r_1,r_2$ respectively.
    We define a new \BR-schema $B$ as follows: first add all the rules of $\hb_1$ to $\hb$ except those of the form $S_\omega,\square\to\bullet$.
    For every rule $S_\omega,\square\to\bullet$ in $\hb_1$, and $S_\sigma,S\to S_{\omega'}$ in $\hb_2$, we define new rules
    \[ S_\omega,S_\sigma \to S_{\omega\sigma},\quad S_{\omega\sigma},S\to S_{\omega\omega'} \]
    and for every other rule of the form $S_{\omega'},S_\sigma\to S_{\omega''}$ in $\hb_2$ we add the new rule $S_{\omega\omega'},S_\sigma\to S_{\omega\omega''}$.
    (We assume that $S_{\omega\sigma}$ and $S_{\omega\omega'}$ have not been used yet, these can of course be replaced with other symbols otherwise.)

    And now for regular expressions of the form $r^+$.
    The language of $r$ has a \BR-schema $B_1$, which has rules of the form $S_\omega,\square\to\bullet$ and $S_\sigma,S_\tau\to S_{\omega'}$.
    We then add the rules $S_\omega,S_\sigma\to S_{\omega\sigma}$ and $S_{\omega\sigma},S_\tau\to S_{\omega'}$.
    \qed

    For example, let us look at the language $\c L(b(a^++aab))$.
    We proceed inductively to generate an initial $\beta$-reducer for this language.

    Firstly, $a^+$ is generated by the rules
    \[ S_a,\square\to\bullet,\quad S_a,S_a\to S_a \]
    And $aab$ is generated by the rules
    \[  S_a,S_a \to S_{aa},\quad S_{aa},S_b\to S_{aab},\quad S_{aab},\square\to\bullet \]

    Thus $a^++aab$ is generated by (only adding the relevant rules):
    \begin{align*}
        S_a,\square&\to\bullet,\\
        S_a,S_a&\to S_{a+aa},\\
        S_{a+aa},S_a&\to S_{\mbox a_1},\\
        S_{a+aa},S_b&\to S_{\mbox{aab}_2},\\
        S_{\mbox a_1},\square&\to\bullet,\\
        S_{\mbox a_1},S_a&\to S_{\mbox a_1},\\
        S_{\mbox{aab}_2},\square&\to\bullet
    \end{align*}

    And finally $b(a^++aab)$ is generated by
    \begin{align*}
        S_b,S_a&\to S_{ba},\\
        S_{ba},\square&\to\bullet,\\
        S_{ba},S_a&\to S_{b(a+aa)},\\
        S_{b(a+aa)},S_a&\to S_{b\mbox a_1},\\
        S_{b(a+aa)},S_b&\to S_{b\mbox{aab}_2},\\
        S_{b\mbox a_1},\square&\to\bullet,\\
        S_{b\mbox a_1},S_a&\to S_{b\mbox a_1},\\
        S_{b\mbox{aab}_2},\square&\to\bullet
    \end{align*}

So we have shown that the class of regular languages, \REG, is a proper subclass of \BR{} (since $\set{a^nb^n}[n\geq1]$ is in \BR{} yet not regular).
The question then becomes, what is the relationship between \BR{} and the class of context-free languages \CFL{}, or the class of deterministic context-free languages \DCFL?

\ari{I hope to prove or disprove this (hopefully finding some analogy for the pumping lemma for \BR)}
We conjecture that $\DCFL=\BR$.
For example, we are unsure but believe that the language of even palindromes, $\set{ww^r}[w\in\set{a,b}^+]$, is not \BR.

\ari{The following section is just me writing down some ideas, far from finished}

\subsection{Multi-$\beta$-Reducible Languages}

An \MBR-schema is a tuple $(\Sigma,\Gamma,s,\set{\hb_i}_{i=0}^n)$ where $\Sigma,\Gamma,s$ are as before and now
\[ \hb_i\colon\Gamma\times(\Gamma\cup\set{\epsilon,\square})\partial(\Gamma\cup\set{\epsilon,\bullet})\times\set{0,\dots,n} \]
For each $i\in\set{0,\dots,n}$ we define $\beta_i$ as before, but instead of returning the string it also returns the index returned by the reduction.
$\beta^*$ will then start using $\beta_0$, and then when the current $\beta_i$ returns $j$, it switches to $\beta_j$.

This is (we believe) stronger than \BR: $\set{ww^r}[w\in\set{a,b}^+]$ is \MBR.
We construct an \MBR-schema as follows:
\[ B = \parens{\set{a,b},\set{A,B,A',B'},[a\to A,b\to B],\set{\hb_0,\hb_1,\hb_2,\hb_3,\hb_4}} \]
Where we have the reduction rules:
\begin{align*}
    \hb_0\colon\quad& A,\epsilon\to A',1\\
        & B,\epsilon\to B',3\\
    \hb_1\colon\quad& A,\square\to A',2\\
    \hb_2\colon\quad& A',A'\to\bullet,2\\
        &A',\epsilon\to\epsilon,2\\
        &A',\square\to A',0\\
    \hb_3\colon\quad& B,\square\to B',2\\
    \hb_4\colon\quad& B',B'\to\bullet,2\\
        &B',\epsilon\to\epsilon,2\\
        &B',\square\to B',0\\
\end{align*}
Intuitively, $\hb_0$ detects the first character and ``tags'' it, then $\hb_1$ or $\hb_3$ ``tag'' the end character if it is the same as the beginning character.
$\hb_2$ and $\hb_4$ ``clean up'' the tags: if the two tagged characters are next to each other, then we can try and accept the word.
Otherwise delete them both.

But \MBR{} is not contained in \CFL{}: $\set{a^nb^nc^n}[n\geq1]\in\MBR$.
Indeed, we can construct an \MBR-schema as follows:
\[ B = \parens{\set{a,b,c},\set{A,B,C,X},[a\to A,b\to B,c\to C],\set{\hb_0,\hb_1,\hb_2,\hb_3}} \]
And we have the reduction rules:
\begin{align*}
    \hb_0\colon\quad& B,A\to X,0\\
        &C,B\to X,0\\
        &C,\square\to C,1\\
    \hb_1\colon\quad& A,\epsilon\to\epsilon,2\\
    \hb_2\colon\quad& B,\epsilon\to\epsilon,3\\
    \hb_3\colon\quad& C,\square\to\bullet,3\\
        &C,\epsilon\to\epsilon,1
\end{align*}
Intuitively, $\hb_0$ checks that $A$ comes before $B$ which comes before $C$ (as otherwise it places in an $X$, which is never reduced).
Then $\hb_1$ matches a single $A$, then calls $\hb_2$ to match a single $B$, which calls $\hb_3$ to match a single $C$.
